---
title: "Population Genetics Homework - Week 12"
author: "Christian Polania"
output: pdf_document
---

```{r setup, include=FALSE}
if (!require("pacman")) install.packages("pacman")
library(pacman)
p_load(tidyverse)
p_load(knitr)
p_load(devtools)
p_load(ggplot2)
p_load(rstanarm)

knitr::opts_chunk$set(echo = TRUE)

greenhouse <- read_csv("/Users/Christian/Desktop/Protea_greenhouse.csv") %>%
  mutate(mom_id = as.factor(paste(species, mom_id, sep = "")))
```

```{r functions, include=FALSE}
heritability <- function(fit, dat) {
  dat_sum <- dat %>%
    group_by(mom_id) %>%
    summarize(count = n())
  n_off <- mean(dat_sum$count)
  df <- as.data.frame(fit)
  sigma_hs <- df$`Sigma[mom_id:(Intercept),(Intercept)]`
  sigma_w <- df$sigma
  h_2 = sigma_hs*n_off/(sigma_hs*n_off + sigma_w^2)
  h_2_df <- tibble(Mean = mean(h_2),
                   lo = quantile(h_2, 0.05),
                   hi = quantile(h_2, 0.95))
  return(h_2_df)
}
```

First, analyses:

```{r analyses}
ven <- subset(greenhouse, species == "venusta")
pun <- subset(greenhouse, species == "punctata")

plma <- stan_glmer(lma ~ (1|mom_id), data = pun, refresh = 0)
pfwc <- stan_glmer(fwc ~ (1|mom_id), data = pun, refresh = 0)
plwr <- stan_glmer(lwr ~ (1|mom_id), data = pun, refresh = 0)
vlma <- stan_glmer(lma ~ (1|mom_id), data = ven, refresh = 0)
vfwc <- stan_glmer(fwc ~ (1|mom_id), data = ven, refresh = 0)
vlwr <- stan_glmer(lwr ~ (1|mom_id), data = ven, refresh = 0)
```

Then a summary of heritabilities:

```{r summaries}
herit <- data.frame(mean = NA, lo = NA, hi = NA)

hplma <- heritability(plma, pun)
hpfwc <- heritability(pfwc, pun)
hplwr <- heritability(plwr, pun)
hvlma <- heritability(vlma, ven)
hvfwc <- heritability(vfwc, ven)
hvlwr <- heritability(vlwr, ven)

herit <-add_row(herit, mean = hplma$Mean, lo = hplma$lo, hi = hplma$hi)
herit <-add_row(herit, mean = hpfwc$Mean, lo = hpfwc$lo, hi = hpfwc$hi)
herit <-add_row(herit, mean = hplwr$Mean, lo = hplwr$lo, hi = hplwr$hi)
herit <-add_row(herit, mean = hvlma$Mean, lo = hvlma$lo, hi = hvlma$hi)
herit <-add_row(herit, mean = hvfwc$Mean, lo = hvfwc$lo, hi = hvfwc$hi)
herit <-add_row(herit, mean = hvlwr$Mean, lo = hvlwr$lo, hi = hvlwr$hi)

herit <- na.omit(herit)
rownames <- c("p.lma", "p.fwc", "p.lwr", "v.lma","v.fwc", "v.lwr")
row.names(herit) <- rownames

herit
````

*Protea punctata* seems to have the same moderate heritability across the board: about a third of phenotypic variation in each of the three studied traits can be explained by genetic variation. *Protea venusta* has lower heritability for leaf mass per area and fresh water content, but has unusually high heritability for leaf length-width ratio.