---
title: "Population Genetics Homework - Week 6"
author: "Christian Polania"
date: "12/17/2021"
output: pdf_document
---

```{r setup, include=FALSE}
if (!require("pacman")) install.packages("pacman")
library(pacman)
p_load(tidyverse)
p_load(knitr)
p_load(devtools)
p_load(tibble)
p_load(hierfstat)

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width=3, fig.height=2)
```

```{r functions, include=FALSE}
infinite_island <- function(n_e, m, mu) {
  return(1/(4*n_e*(m + mu) + 1))
}

lambda <- function(mu, m, k) {
  m_prime <- m*k/(k - 1)
  l <- (1/2)*((1 - mu)^(-2)*(1- m_prime)^(-2) - 1)
  return(l)
}

gamma <- function(mu, m, k) {
  m_prime <- m*k/(k - 1)
  g <- (1 + m_prime*(2 - m_prime)/(k*mu*(2 - mu)*(1 - m_prime)^2))^(-1)
  return(g)
}

finite_island <- function(n_e, m, mu, n_islands) {
  l <- lambda(mu, m, n_islands)
  g <- gamma(mu, m, n_islands)
  h_s <- 4*n_e*l*g/(4*n_e*l*g + 1)
  h_b <- h_s*(1 + 4*n_e*l)/(4*n_e*l)
  h_t <- ((n_islands - 1)/n_islands)*h_b + (1/n_islands)*h_s
  f_st <- (h_t - h_s)/h_t
  return(f_st)
}

n_e <- 100
mu <- 10^(-4)
k_vals <- rep(c(5, 25, 250, 2500), 2)
m_vals <- c(rep(0.01, 4), rep(0.05, 4))
dat <- data.frame(k = k_vals,
                  m = m_vals,
                  Infinite = rep(NA, length(k_vals)),
                  Finite = rep(NA, length(m_vals)),
                  Ratio = rep(NA, length(k_vals)))
for (i in 1:nrow(dat)) {
  dat$Infinite[i] <- infinite_island(n_e, dat$m[i], mu)
  dat$Finite[i] <- finite_island(n_e, dat$m[i], mu, dat$k[i])
  dat$Ratio[i] <- dat$Infinite[i]/dat$Finite[i]
}
round(dat, 3)

initialize <- function(k, n_loci) {
  p <- array(dim = c(n_loci, k, 2))
  for (j in 1:n_loci) {
    for (i in 1:k) {
      p[j, i, 1] <- runif(n = 1, min = 0, max = 1)
      p[j, i, 2] <- 1 - p[j, i, 1]
    }
  }
  return(p)
}

make_symmetric_matrix <- function(m, k) {
  M <- diag(x = 1 - m, nrow = k)
  for (i in 1:(k - 1)) {
    for (j in (i+1):k) {
      M[i, j] <- m/(k-1)
      M[j, i] <- M[i, j]
    }
  }
  return(M)
}

sample_two_alleles <- function(p) {
  allele_1 <- ((runif(1) < p) + 1)*10
  allele_2 <- (runif(1) < p) + 1
  return(allele_1 + allele_2)
}

make_sample <- function(p, n_sample) {
  n_pops <- dim(p)[2]
  n_loci <- dim(p)[1]
  names <- paste("Locus", seq(1:n_loci), sep = "")
  pops <- numeric(n_pops*n_sample)
  ct <- 1
  for (k in 1:n_pops) {
    population <- paste("Population", k, sep = "")
    for (i in 1:n_sample) {
      pops[ct] <- population
      ct <- ct + 1
    }
  }
  locus_data <- matrix(nrow = n_pops*n_sample, ncol = n_loci)
  for (i in 1:n_loci) {
   ct <- 1
    for (k in 1:n_pops) {
      for (n in 1:n_sample) {
        locus_data[ct, i] <- sample_two_alleles(p[i, k, 1])
        ct <- ct + 1
      }
    }
  }
  df <- data.frame(locus_data)
  colnames(df) <- paste("Locus", seq(1:n_loci), sep = "")
  df$Population <- pops
  df <- relocate(df, Population)
  return(df)
}

simulate <- function(n_e, m, mu, n_islands, n_loci, n_gen, n_sample) {
  p <- initialize(n_islands, n_loci)
  M <- make_symmetric_matrix(m, n_islands)
  V <- make_symmetric_matrix(mu, 2)
  g_st <- numeric(n_loci)
  p_star <- array(dim = c(n_loci, n_islands, 2))
  for (i in 1:n_gen) {
    for (k in 1:n_loci) {
      p_star[k, , ] <- M %*% p[k, , ] %*% V
      for (j in 1:n_islands) {
        p[k, j, ] <- rmultinom(1, 2*n_e, p_star[k, j, ])/(2*n_e)
      }
      mu_p <- mean(p[k, , 1])
      g_st[k] <- ((n_islands - 1)/n_islands)*var(p[k, , 1])/(mu_p*(1 - mu_p))
    }
  }
  df <- make_sample(p, n_sample)
  return(wc(df)$FST)
}

run_simulation <- function(n_e, m, mu, n_islands, n_loci, n_gen, 
                           n_sample, n_repetitions)
{
  f_st <- numeric(n_repetitions)
  for (i in 1:n_repetitions) {
    if (n_repetitions > 50) {
      cat(".", sep="")
      if ((i %% 50) == 0) {
        cat(i, "\n", sep ="")
      }
    }
    f_st[i] <-simulate(n_e, m, mu, n_islands, n_loci, n_gen, n_sample)
  }
  return(f_st)
}

plot_simulation <- function(f_st, n_e, m, mu, n_islands) {
  df <- data.frame(F_st = f_st)
  p <- ggplot(df, aes(x = F_st)) +
    geom_histogram(bins = 20, alpha = 0.4) +
    geom_vline(xintercept = infinite_island(n_e, m, mu),
               linetype = "dashed",
               color = "red") +
    geom_vline(xintercept = finite_island(n_e, m, mu, n_islands),
               linetype = "dashed",
               color = "blue") +
    ggtitle(paste("N_e = ", n_e, ", m = ", m, ", mu = ", mu, 
                  ", n_islands = ", n_islands, sep = "")) +
    theme_bw()
  return(p)
}
```

**Intro**  
  
I ran 7 sets of simulations to explore the variability generated by, and the usefulness of, the infinite and finite island approximations. Starting with the original parameters to be used as a middle ground, I then ran an "over" and "under" set of simulations using a low and high value for n_loci, n_sample, and n_gen. I'll show the low, middle and high value simulations for each, with two plots and two model summaries (one plot and summary each for infinite and finite). Then, I'll discuss.

```{r 0: 10-25-1000, include=FALSE}
n_loci <- 10
n_sample <- 25
n_gen <- 1000
n_repetitions <- 10

df <- data.frame(N_e = NA, m = NA, mu = NA, n_islands = NA, Observed = NA,
                 Infinite = NA, Finite = NA)
for (n_e in c(25, 100, 250, 500)) {
  for (m in c(0.05, 0.01, 0.005, 0.001)) {
    for (mu in c(0.001, 1e-4)) {
      for (n_islands in c(10, 25, 50)) {
        cat("N_e = ", n_e, ", m = ", m, ", mu = ", mu, 
            "n_islands = ", n_islands, "\n" )
        f_st <- run_simulation(n_e, m, mu, n_islands, n_loci, n_gen, 
                               n_sample, n_repetitions)
        df <- add_row(df, 
                      N_e = rep(n_e, n_repetitions),
                      m = rep(m, n_repetitions), 
                      mu = rep(mu, n_repetitions), 
                      n_islands = rep(n_islands, n_repetitions),
                      Observed = f_st,
                      Infinite = infinite_island(n_e, m, mu),
                      Finite = finite_island(n_e, m, mu, n_islands))
      }
    }
  }
}
df <- filter(df, !is.na(N_e))


p0a <- ggplot(df, aes(x = Infinite, y = Observed)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", 
              color = "red") + theme_bw() +
  ggtitle("infinite, 10-25-1000")

p0b <- ggplot(df, aes(x = Finite, y = Observed)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", 
              color = "red") + theme_bw() +
  ggtitle("finite, 10-25-1000")

s0a <- summary(lm(Observed ~ Infinite, data = df))
s0b <- summary(lm(Observed ~ Finite, data = df))
```

```{r 1: 20-25-1000, include=FALSE}
n_loci <- 20
n_sample <- 25
n_gen <- 1000
n_repetitions <- 10

df <- data.frame(N_e = NA, m = NA, mu = NA, n_islands = NA, Observed = NA,
                 Infinite = NA, Finite = NA)
for (n_e in c(25, 100, 250, 500)) {
  for (m in c(0.05, 0.01, 0.005, 0.001)) {
    for (mu in c(0.001, 1e-4)) {
      for (n_islands in c(10, 25, 50)) {
        cat("N_e = ", n_e, ", m = ", m, ", mu = ", mu, 
            "n_islands = ", n_islands, "\n" )
        f_st <- run_simulation(n_e, m, mu, n_islands, n_loci, n_gen, 
                               n_sample, n_repetitions)
        df <- add_row(df, 
                      N_e = rep(n_e, n_repetitions),
                      m = rep(m, n_repetitions), 
                      mu = rep(mu, n_repetitions), 
                      n_islands = rep(n_islands, n_repetitions),
                      Observed = f_st,
                      Infinite = infinite_island(n_e, m, mu),
                      Finite = finite_island(n_e, m, mu, n_islands))
      }
    }
  }
}
df <- filter(df, !is.na(N_e))


p1a <- ggplot(df, aes(x = Infinite, y = Observed)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", 
              color = "red") + theme_bw() +
  ggtitle("infinite, 20-25-1000")

p1b <- ggplot(df, aes(x = Finite, y = Observed)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", 
              color = "red") + theme_bw() +
  ggtitle("finite, 20-25-1000")

s1a <- summary(lm(Observed ~ Infinite, data = df))
s1b <- summary(lm(Observed ~ Finite, data = df))
```

```{r 2: 10-50-1000, include=FALSE}
n_loci <- 10
n_sample <- 50
n_gen <- 1000
n_repetitions <- 10

df <- data.frame(N_e = NA, m = NA, mu = NA, n_islands = NA, Observed = NA,
                 Infinite = NA, Finite = NA)
for (n_e in c(25, 100, 250, 500)) {
  for (m in c(0.05, 0.01, 0.005, 0.001)) {
    for (mu in c(0.001, 1e-4)) {
      for (n_islands in c(10, 25, 50)) {
        cat("N_e = ", n_e, ", m = ", m, ", mu = ", mu, 
            "n_islands = ", n_islands, "\n" )
        f_st <- run_simulation(n_e, m, mu, n_islands, n_loci, n_gen, 
                               n_sample, n_repetitions)
        df <- add_row(df, 
                      N_e = rep(n_e, n_repetitions),
                      m = rep(m, n_repetitions), 
                      mu = rep(mu, n_repetitions), 
                      n_islands = rep(n_islands, n_repetitions),
                      Observed = f_st,
                      Infinite = infinite_island(n_e, m, mu),
                      Finite = finite_island(n_e, m, mu, n_islands))
      }
    }
  }
}
df <- filter(df, !is.na(N_e))


p2a <- ggplot(df, aes(x = Infinite, y = Observed)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", 
              color = "red") + theme_bw() +
  ggtitle("infinite, 10-50-1000")

p2b <- ggplot(df, aes(x = Finite, y = Observed)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", 
              color = "red") + theme_bw() +
  ggtitle("finite, 10-50-1000")

s2a <- summary(lm(Observed ~ Infinite, data = df))
s2b <- summary(lm(Observed ~ Finite, data = df))
```

```{r 3: 10-25-2000, include=FALSE}
n_loci <- 10
n_sample <- 25
n_gen <- 2000
n_repetitions <- 10

df <- data.frame(N_e = NA, m = NA, mu = NA, n_islands = NA, Observed = NA,
                 Infinite = NA, Finite = NA)
for (n_e in c(25, 100, 250, 500)) {
  for (m in c(0.05, 0.01, 0.005, 0.001)) {
    for (mu in c(0.001, 1e-4)) {
      for (n_islands in c(10, 25, 50)) {
        cat("N_e = ", n_e, ", m = ", m, ", mu = ", mu, 
            "n_islands = ", n_islands, "\n" )
        f_st <- run_simulation(n_e, m, mu, n_islands, n_loci, n_gen, 
                               n_sample, n_repetitions)
        df <- add_row(df, 
                      N_e = rep(n_e, n_repetitions),
                      m = rep(m, n_repetitions), 
                      mu = rep(mu, n_repetitions), 
                      n_islands = rep(n_islands, n_repetitions),
                      Observed = f_st,
                      Infinite = infinite_island(n_e, m, mu),
                      Finite = finite_island(n_e, m, mu, n_islands))
      }
    }
  }
}
df <- filter(df, !is.na(N_e))


p3a <- ggplot(df, aes(x = Infinite, y = Observed)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", 
              color = "red") + theme_bw() +
  ggtitle("infinite, 10-25-2000")

p3b <- ggplot(df, aes(x = Finite, y = Observed)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", 
              color = "red") + theme_bw() +
  ggtitle("finite, 10-25-2000")

s3a <- summary(lm(Observed ~ Infinite, data = df))
s3b <- summary(lm(Observed ~ Finite, data = df))
```

```{r 4: 5-25-1000, include=FALSE}
n_loci <- 5
n_sample <- 25
n_gen <- 1000
n_repetitions <- 10

df <- data.frame(N_e = NA, m = NA, mu = NA, n_islands = NA, Observed = NA,
                 Infinite = NA, Finite = NA)
for (n_e in c(25, 100, 250, 500)) {
  for (m in c(0.05, 0.01, 0.005, 0.001)) {
    for (mu in c(0.001, 1e-4)) {
      for (n_islands in c(10, 25, 50)) {
        cat("N_e = ", n_e, ", m = ", m, ", mu = ", mu, 
            "n_islands = ", n_islands, "\n" )
        f_st <- run_simulation(n_e, m, mu, n_islands, n_loci, n_gen, 
                               n_sample, n_repetitions)
        df <- add_row(df, 
                      N_e = rep(n_e, n_repetitions),
                      m = rep(m, n_repetitions), 
                      mu = rep(mu, n_repetitions), 
                      n_islands = rep(n_islands, n_repetitions),
                      Observed = f_st,
                      Infinite = infinite_island(n_e, m, mu),
                      Finite = finite_island(n_e, m, mu, n_islands))
      }
    }
  }
}
df <- filter(df, !is.na(N_e))


p4a <- ggplot(df, aes(x = Infinite, y = Observed)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", 
              color = "red") + theme_bw() +
  ggtitle("infinite, 5-25-1000")

p4b <- ggplot(df, aes(x = Finite, y = Observed)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", 
              color = "red") + theme_bw() +
  ggtitle("finite, 5-25-1000")

s4a <- summary(lm(Observed ~ Infinite, data = df))
s4b <- summary(lm(Observed ~ Finite, data = df))
```

```{r 5: 10-5-1000, include=FALSE}
n_loci <- 10
n_sample <- 5
n_gen <- 1000
n_repetitions <- 10

df <- data.frame(N_e = NA, m = NA, mu = NA, n_islands = NA, Observed = NA,
                 Infinite = NA, Finite = NA)
for (n_e in c(25, 100, 250, 500)) {
  for (m in c(0.05, 0.01, 0.005, 0.001)) {
    for (mu in c(0.001, 1e-4)) {
      for (n_islands in c(10, 25, 50)) {
        cat("N_e = ", n_e, ", m = ", m, ", mu = ", mu, 
            "n_islands = ", n_islands, "\n" )
        f_st <- run_simulation(n_e, m, mu, n_islands, n_loci, n_gen, 
                               n_sample, n_repetitions)
        df <- add_row(df, 
                      N_e = rep(n_e, n_repetitions),
                      m = rep(m, n_repetitions), 
                      mu = rep(mu, n_repetitions), 
                      n_islands = rep(n_islands, n_repetitions),
                      Observed = f_st,
                      Infinite = infinite_island(n_e, m, mu),
                      Finite = finite_island(n_e, m, mu, n_islands))
      }
    }
  }
}
df <- filter(df, !is.na(N_e))


p5a <- ggplot(df, aes(x = Infinite, y = Observed)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", 
              color = "red") + theme_bw() +
  ggtitle("infinite, 10-5-1000")

p5b <- ggplot(df, aes(x = Finite, y = Observed)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", 
              color = "red") + theme_bw() +
  ggtitle("finite, 10-5-1000")

s5a <- summary(lm(Observed ~ Infinite, data = df))
s5b <- summary(lm(Observed ~ Finite, data = df))
```

```{r 6: 10-25-200, include=FALSE}
n_loci <- 10
n_sample <- 25
n_gen <- 200
n_repetitions <- 10

df <- data.frame(N_e = NA, m = NA, mu = NA, n_islands = NA, Observed = NA,
                 Infinite = NA, Finite = NA)
for (n_e in c(25, 100, 250, 500)) {
  for (m in c(0.05, 0.01, 0.005, 0.001)) {
    for (mu in c(0.001, 1e-4)) {
      for (n_islands in c(10, 25, 50)) {
        cat("N_e = ", n_e, ", m = ", m, ", mu = ", mu, 
            "n_islands = ", n_islands, "\n" )
        f_st <- run_simulation(n_e, m, mu, n_islands, n_loci, n_gen, 
                               n_sample, n_repetitions)
        df <- add_row(df, 
                      N_e = rep(n_e, n_repetitions),
                      m = rep(m, n_repetitions), 
                      mu = rep(mu, n_repetitions), 
                      n_islands = rep(n_islands, n_repetitions),
                      Observed = f_st,
                      Infinite = infinite_island(n_e, m, mu),
                      Finite = finite_island(n_e, m, mu, n_islands))
      }
    }
  }
}
df <- filter(df, !is.na(N_e))


p6a <- ggplot(df, aes(x = Infinite, y = Observed)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", 
              color = "red") + theme_bw() +
  ggtitle("infinite, 10-25-200")

p6b <- ggplot(df, aes(x = Finite, y = Observed)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", 
              color = "red") + theme_bw() +
  ggtitle("finite, 10-25-200")

s6a <- summary(lm(Observed ~ Infinite, data = df))
s6b <- summary(lm(Observed ~ Finite, data = df))
```
  
  
**Effect of number of number of loci (2, 10, 20)**  
```{r loci, echo=FALSE}
p4a
p4b
s4a
s4b
p0a
p0b
s0a
s0b
p1a
p1b
s1a
s1b
```
  
**Effect of number of sample size (2, 10, 20)**  
```{r samp, echo=FALSE}
p5a
p5b
s5a
s5b
p0a
p0b
s0a
s0b
p2a
p2b
s2a
s2b
```
  
**Effect of number of number of generations (2, 10, 20)**  
```{r gen, echo=FALSE}
p6a
p6b
s6a
s6b
p0a
p0b
s0a
s0b
p3a
p3b
s3a
s3b
```
  
**Analysis**  
  
*How close is the relationship between observed estimates of FST from the simulations and those predicted from the infinite island model? from the finite island model?*
The strength of the relationship between the observed FSTs and the estimated FSTs were consistently higher in the finite model. Higher loci number, sample size, and generation number each improved the fit, but it was always the same amount for finite and infinite.  
  
*Under what conditions is the observed variation in FST relatively small?*
For all simulations, Fst estimates grew less variable with higher parameters except for changes in generation number. Interestingly, this improvement was consistently larger for the infinite models.  
  
*Are there conditions under which the predictions of the infinite island and finite island model are similar enough that we don’t need to worry about the added mathematical complexity of the finite island model?*
As FST estimates grew more consistent with higher parameters and this difference was larger for infinite models, the simulations where the models performed most similarly were with lower number of loci and lower sample size. 


```{r, include=FALSE}
options(tinytex.verbose = TRUE)
```